---
title: "01 - Read data, preprocessing, and QC"
---

## Load libraries

```{r, message=FALSE}
# Load libraries for spatial transcriptomics processing, plotting, and data manipulation.
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Prepare output folders

```{r}
# These folders store the plots and Seurat objects created by this lesson.
dir.create("../output/plots/01_Data_processing_and_QC", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/spatial_rds_task", recursive = TRUE, showWarnings = FALSE)
```

## Load data

Public data have been downloaded from 10x Genomics (https://www.10xgenomics.com/datasets) and stored in the `../data/visium/BCLL_10_T` folder.
The data is in the standard 10x Genomics format, which includes a `filtered_feature_bc_matrix` folder and a `spatial` folder with the tissue image and spatial coordinates.

```{r}
# Load the 10x Visium dataset.
seurat_object <- Load10X_Spatial("../data/visium/BCLL_10_T")

# Adjust the low-resolution scale factor to match the high-resolution scale factor.
# This helps the tissue image and spots align correctly for this dataset.
seurat_object@images[["slice1"]]@scale.factors[["lowres"]] <- seurat_object@images[["slice1"]]@scale.factors[["hires"]]
```

------------------------------------

# QC and data preprocessing
## Get QC metrics

```{r}
# Regex variables.
# ##################################################
# Define regex patterns to identify mitochondrial and ribosomal genes.
mt_regex <- "^MT-"
ribo_regex <- "^RP[SL]"
```


```{r}
# Calculate mitochondrial content.
# ##################################################
# Calculate the percentage of mitochondrial gene expression per spot for QC.
seurat_object <- PercentageFeatureSet(seurat_object, pattern = mt_regex, col.name = "percent_mt")

# Calculate ribosomal content.
# ##################################################
# Calculate the percentage of ribosomal gene expression per spot for QC.
seurat_object <- PercentageFeatureSet(seurat_object, pattern = ribo_regex, col.name = "percent_ribo")

```

## TASK 1: Identify informative gene sets

We have used **mitochondrial genes** and **ribosomal genes** for quality control and visualization. Now it is your turn to think beyond them.

## Plot QC metrics

```{r}
# Generate QC metric plots.
# ##################################################
# Visualize total counts, number of detected genes, and percentage of mitochondrial/ribosomal transcripts.
# ----------
# nCount_Spatial plot
# ----------
plot1 <- VlnPlot(seurat_object, features = "nCount_Spatial", pt.size = 0.1) +
         NoLegend() + theme_bw() + theme(legend.position = "none")

# ----------
# nFeature_Spatial plot
# ----------
plot2 <- VlnPlot(...)

# ----------
# percent_mt plot
# ----------
plot3 <- VlnPlot(...)

# ----------
# percent_ribo plot
# ----------
plot4 <- VlnPlot(...)


# ##################################################
# Arrange all plots.
(plot1 | plot2 | plot3 | plot4)

```

## Perform QC

```{r, warning=FALSE}
# Remove non-informative genes.
# ##################################################
# Remove genes used for QC to avoid biasing downstream analysis.

# Combine regex patterns with "|" to mean "or".
regex_remove_genes <- paste0(mt_regex, "|", ribo_regex)

# Extract all genes to exclude.
remove_me <- rownames(seurat_object)[grepl(regex_remove_genes, rownames(seurat_object))]

# Keep all genes that are not in remove_me.
seurat_object <- subset(seurat_object, features = rownames(seurat_object)[!rownames(seurat_object) %in% remove_me])

# ##################################################
# Subset spots from the Seurat object based on selected thresholds.
nFeature_Spatial_threshold_lower <- 300
nCount_Spatial_threshold_lower <- 300
percent_mt_threshold <- 5

seurat_object_size_before <- dim(seurat_object)


# Subset the Seurat object based on the thresholds defined above.
seurat_object <- subset(seurat_object,
                        subset = ...
                        )

```

Compare the number of spots before and after QC filtering.
```{r, warning=FALSE}
seurat_object_size_after <- ...


```


## Spatial feature plots

```{r}
# Create spatial plots to inspect QC metrics.
# ##################################################
# Visualize the QC metrics in a spatial context.
p <- SpatialFeaturePlot(seurat_object, features = c("nCount_Spatial"))
print(p)

ggsave("../output/plots/01_Data_processing_and_QC/spatial_qc_metrics.png", plot = p, width = 10, height = 8)
```

Check other features like nFeature_Spatial, percent_mt...
```{r}
# Create spatial plots to inspect QC metrics.
# ##################################################
# Visualize the QC metrics in a spatial context.
p <- SpatialFeaturePlot(...)
print(p)
```

### Some additional steps to make life easier later on

```{r}
# ##################################################
# Fix the layer dimension names. This makes later Seurat v5 steps more reliable.
seurat_object@assays[["Spatial"]]@layers[["counts"]]@Dimnames[[2]] <- colnames(seurat_object@assays[["Spatial"]])
seurat_object@assays[["Spatial"]]@layers[["counts"]]@Dimnames[[1]] <- rownames(seurat_object@assays[["Spatial"]])

# ##################################################
# Save the Seurat object for lesson 2.
saveRDS(seurat_object, "../data/spatial_rds_task/1_tonsil_qc_performed.rds")
```

## Handoff to lesson 2

In the next lesson, we will load `../data/spatial_rds_task/1_tonsil_qc_performed.rds` and use it for normalization, PCA, clustering, and spatial cluster visualization.

