---
title: "02 - Clustering"
---

## Load libraries

```{r, message=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Prepare output folders

```{r}
dir.create("../output/plots/02_Clustering", recursive = TRUE, showWarnings = FALSE)
dir.create("../data/spatial_rds_task", recursive = TRUE, showWarnings = FALSE)
```

## Load the QC-filtered object

```{r}
# This object was created at the end of lesson 1.
seurat_object <- readRDS("../data/spatial_rds_task/1_tonsil_qc_performed.rds")

# Work with the spatial expression assay.
DefaultAssay(seurat_object) <- "Spatial"
```

## Create a new assay

```{r}
# Transform the raw counts using log1p and wrap them into a new Seurat assay for downstream use.

# ##################################################
# Get the log1p data
log1p_data <- seurat_object@assays[["Spatial"]]@layers[["counts"]] %>% log1p()

# ##################################################
# Create assay object
log1p_assay <- CreateAssayObject(data = log1p_data)
```

```{r, warning=FALSE}
# Add the new log1p assay to the object and set it as default for upcoming steps.
# ##################################################
# Create new seurat assay from the log1p data
seurat_object[["log1p"]] <- log1p_assay

# Set the default assay to the "log1p" for the pca calculation
DefaultAssay(seurat_object) <- "log1p"
```

## Scale the data

```{r}
# Scale the data to normalize variance across genes before PCA.
# ##################################################
# Define all gene names as variables. NOTE: In some tutorials you will see the usage of Highly variable genes only.
all_genes <- rownames(seurat_object)

# Scale the data prior to dimensional reduction (eg. PCA)
seurat_object <- ScaleData(seurat_object, features = all_genes)
```

## Dimensionality reduction, clustering, and visualization

```{r}
# Run PCA on log1p-transformed data to reduce dimensionality and condense information.
# ##################################################
# Run PCA
seurat_object <- RunPCA(seurat_object, assay = "log1p", verbose = FALSE, features = rownames(seurat_object), npcs=50)
```


## Inspect PCA

```{r}
# The elbow plot helps us choose how many PCs to use for clustering.
p_elbow <- ElbowPlot(seurat_object, ndims = 30)
p_elbow

ggsave("../output/plots/02_Clustering/elbow_plot.png", plot = p_elbow, width = 6, height = 4)
```

## Find clusters

```{r}
# Use the first 20 PCs for this beginner tutorial.
# The resolution controls how fine or coarse the clusters are.
pcs_to_use <- 1:20
cluster_resolution <- 0.8

seurat_object <- FindNeighbors(seurat_object, dims = pcs_to_use)
seurat_object <- FindClusters(seurat_object,
  resolution = cluster_resolution,
  algorithm = 2)
```

Remark: In practice we would recommend using leiden (algorithm = 4) clustering instead of louvain (algorithm = 2).
Because Leiden requires extra python package installation.
Therefore we stick to Louvain for the sake of simplicity in this workshop.

## TASK 2: Try another clustering resolution

Change `cluster_resolution` to `0.2` or `0.8`, rerun the clustering step, and compare the number of clusters.

Questions:

- What happens when the resolution is lower?
- What happens when the resolution is higher?
- Which resolution gives clusters that look spatially coherent?

## Run UMAP

```{r}
# UMAP places similar spots close together in a two-dimensional plot.
seurat_object <- RunUMAP(seurat_object, dims = pcs_to_use)
```

## Visualize clusters

```{r}
# UMAP view: spots close together have similar expression profiles.
p_umap <- DimPlot(seurat_object, reduction = "umap", group.by = "seurat_clusters", label = TRUE) +
  NoLegend()
p_umap

ggsave("../output/plots/02_Clustering/umap_clusters.png", plot = p_umap, width = 6, height = 5)
```

```{r}
# Spatial view: the same clusters shown on the tissue image.
p_spatial_clusters <- SpatialDimPlot(seurat_object, group.by = "seurat_clusters", label = TRUE, label.size = 3)
p_spatial_clusters

ggsave("../output/plots/02_Clustering/spatial_clusters.png", plot = p_spatial_clusters, width = 8, height = 7)
```

## Save the clustered object

```{r}
# Save this object for marker analysis in lesson 3.
saveRDS(seurat_object, "../data/spatial_rds_task/2_tonsil_clustered.rds")
```

## Handoff to lesson 3

In the next lesson, we will load `../data/spatial_rds_task/2_tonsil_clustered.rds` and find marker genes for each cluster.
