---
title: "03 - Marker analysis"
---

## Load libraries

```{r, message=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Prepare output folders

```{r}
dir.create("../output/plots/03_Marker_analysis", recursive = TRUE, showWarnings = FALSE)
dir.create("../output/tables", recursive = TRUE, showWarnings = FALSE)
```

## Load the clustered object

```{r}
# This object was created at the end of lesson 2.
seurat_object <- readRDS("../data/spatial_rds_task/2_tonsil_clustered.rds")

DefaultAssay(seurat_object) <- "Spatial"
Idents(seurat_object) <- "seurat_clusters"
```

## Find cluster marker genes

```{r}
# Find genes that are enriched in each cluster compared with all other clusters.
cluster_markers <- FindAllMarkers(
  seurat_object,
  only.pos = TRUE
)

head(cluster_markers)
```

Check logfc.threshold and min.pct parameters in the `FindAllMarkers()` function to adjust the stringency of marker detection.
Run the function with different parameters to see how the number of detected markers changes.


## Save marker table

```{r}
# Save the full marker table for later inspection.
write.csv(cluster_markers, "../output/tables/03_cluster_markers.csv", row.names = FALSE)
```

## Inspect top markers

```{r}
# Select the top 5 marker genes per cluster.
top_markers <- cluster_markers %>%
  group_by(cluster) %>%
  slice_max(n = 5, order_by = avg_log2FC) %>%
  ungroup()

top_markers
```

```{r}
# Pick a few top marker genes to visualize.
marker_genes <- unique(top_markers$gene)
marker_genes <- marker_genes[1:min(4, length(marker_genes))]

marker_genes
```

## Visualize marker expression

```{r}
# UMAP expression view for selected marker genes.
p_feature <- FeaturePlot(seurat_object, features = marker_genes, reduction = "umap", ncol = 2)
p_feature

ggsave("../output/plots/03_Marker_analysis/umap_marker_features.png", plot = p_feature, width = 10, height = 8)
```

```{r}
# Violin plots compare expression across clusters.
p_violin <- VlnPlot(seurat_object, features = marker_genes, pt.size = 0, ncol = 2)
p_violin

ggsave("../output/plots/03_Marker_analysis/violin_marker_features.png", plot = p_violin, width = 10, height = 8)
```

```{r}
# Spatial plots show where marker genes are expressed in the tissue.
p_spatial_markers <- SpatialFeaturePlot(seurat_object, features = marker_genes, ncol = 2)
p_spatial_markers

ggsave("../output/plots/03_Marker_analysis/spatial_marker_features.png", plot = p_spatial_markers, width = 10, height = 8)
```

## TASK 3: Interpret one cluster

Choose one cluster and inspect its top marker genes.

Questions:

- Which genes are most enriched in your chosen cluster?
- Where is the cluster located in the tissue?
- Do the marker genes suggest a possible biological identity or tissue region?

Choose one marker gene from the table and make your own `SpatialFeaturePlot()` for it.


## Optional task

Instead of comparing one cluster to all other clusters, you can also compare two specific clusters. Use the `FindMarkers()` function to do this.
Choose 2 clusters and compare them. What are the top marker genes that distinguish these two clusters?

```{r}
FindMarkers(...)
```